### 2012-10-15 rnaseq_paper.tex
merged with Titus's update

We next used Tophat to align mRNAseq reads to the genome and partition
reads by chromosome; we then assembled the partitioned reads with
Velvet and Oases using the same range of parameters as the global
assembly, above.  While the local assemblies were considerably more
computationally efficient, they lacked several thousand unique regions
that were present in the global assembly
(Table~\ref{unique_sequences_matched_mouse}); this is probably due to
the incomplete nature of the current chicken genome (cite).
Interestingly, over a hundred regions were present {\em only} in the
local assemblies, suggesting that the local assemblies might be
recovering additional exons.  Significant numbers of unique regions
from both global and local assemblies showed homology to the mouse
genome, indicating that at least some of these unique sequences
represented real sequence content.  Figure~\ref{global_vs_local} shows
an example of different isoforms detected by the two assembly methods.

@@

We developed an exon graph approach to construct gene models from
alignments of transcripts against the genome.  Our approach, called
Gimme, uses exon graphs to merge transcripts and gene models based on
overlapping exons (see Materials and Methods).  We used Gimme to
obtain 20,439 gene models containing 25,859 isoforms from our global
assembly; 21,963 gene models containing 25,859 XX from our local
assembly; and 20,922 gene models containing 27,732 isoforms from the
merged global and local assemblies (Table~\ref{genes_transcripts}).

(more here - what? @CTB)

@@

\subsubsection{Most splice junctions are independently supported}

@@

As shown above, Cufflinks and the assembly method detect a number of
distinct but equally valid splice junctions, which suggests that we
could obtain greater sensitivity by merging both sets of predictions.
We therefore used Gimme to merge the Cufflinks and assembly gene
models Table.~\ref{genes_transcripts}.  This resulted in a decreased
number of total genes, suggesting that some fragmented genes were
merged together to form more complete gene structures (e.g. see
Fig.~\ref{cufflinks_asm}).
% @CTB what about splice junctions etc??

@@

\subsection{Leftover text}

The program can merge all transcripts from multiple-kmers and predict a structure of full-length isoforms and genes.
As shown in figure~\ref{model_comparisons}, our gene models are comparable to other existing gene models.
Total of 14,703 genes were obtained from global assembly and 14,931 genes from local assembly.
However, the number of genes obtained from global + local assembly is only 15,411.
The number of genes and isoforms slightly increased after combining transcripts from global and local
assembly together (Table~\ref{genes_transcripts}). 
This indicates that some transcripts from both methods were merged together.

Most transcripts include a large part of untranslated regions (UTRs), especially $3'$ UTR in our datasets.
These regions are challenging to predict correctly solely from computational methods due to low degree of conservation [?].
RNA-Seq-derived gene models are useful for studying variations within UTR regions, which may be involved in regulation of isoform expression\cite{} (Fig.~\ref{long_utr}).

@@

The number of genes from our pipeline might be overestimated due to
fragmented transcripts.  In our datasets, read coverage is lower at
the $5'$ end due to the method used to convert mRNA to cDNA in library
preparation.  A bias of read coverage toward $3'$ end and low
expression level result in fragmented transcripts as shown in
Fig.~\ref{fragmented_transcripts}.  Moreover, spurious splice
junctions due to duplication in genome sequence can lead to a large
number of dubious isoforms.  This seems to occur in a few genome
sequences such as chromosome E64\_random.
Figure~\ref{isoforms_boxplot} shows that a majority of genes from our
gene models have only a few isoforms.

@@

\subsubsection{Cufflinks}

To evaluate an efficiency of our pipeline in detecting splice junctions,
we compared the number of splice junctions detected by our pipeline to that from , ESTs and
Ensembl gene models version 64.
Cufflinks detected 89,289 (86.49\%) splice junctions that are also detected by our pipeline.
However, 37.61\% (5,565) of junctions not detected by Cufflinks are supported by ESTs.
This suggests that the assembly detects some splice junctions that are not detected by Cufflinks and they are
genuine splice junctions.

We also aligned transcripts from Cufflinks and the pipeline to Ensembl transcripts to compare genes and transcripts detected by
both methods.
We selected only transcripts that match with more than 90\% identity and cover more than 80\% of an Ensembl transcript.
Our pipeline detects 7,046 genes and Cufflinks detects 8,341 genes of total 17,934 Ensembl genes.
6,572 of those genes are both detectd by our pipeline and Cufflinks.
In addition, we wanted to compare the number of isoforms detected by our pipeline and Cufflinks within the same gene; however,
a transcript can match to multiple isoforms in the same gene, especially when it is not complete.
Therefore, we compare the number of splice junctions detected by both methods within the same gene instead.
From 6,572 genes, our pipeline detects 57,437 splice junctions, whereas Cufflinks detects 62,101 junctions.

However, because Ensembl gene models do not include all isoforms or splice junctions from ESTs, we investigated splice junctions
that are not in Ensembl but are supported by ESTs in those 6,572 genes and found that our pipeline detects 2,260 splice junctions
and Cufflinks only detects 1,588 splice junctions. Figure ~\ref{alt_splice_site} shows an example of splice junctions not included in Ensembl gene models, which are detected by our pipeline.

@@

Single and paired-end reads were mapped to chicken genome by Tophat2\cite{Trapnell:2009dp} release 2.0.0 using
default parameters without annotations.
All reads were mapped to cDNA sequences derived from gene models by Bowtie2\cite{Langmead:2009fv} with
default parameters \texttt{(n=2, l=28, e=70, k=1)}.

@@

\caption{
\bf{Number of total and unique splice junctions}}
\begin{tabular}{cccccc}
\hline
Method& Total & Unique & Supported by ESTs \\ 
\hline
Oases-M assembly & 114,336 & 8,735 & 447 (5.1\%) \\
Unmerged assembly & 115,159 & 9,558 & 1,640 (17.15\%) \\
\hline
%table information
\end{tabular}
%\begin{flushleft}\footnotesize \textit{*Minumum set}
%\end{flushleft}
\label{Oases-M}

\caption{
\bf{Number of putative genes and isoforms}}
\begin{tabular}{cccccc}
\hline
Method& Gene & Isoform \\ 
\hline
Global & 14,703 & 28,406 \\
Local & 14,931 & 22,127 \\
Global + Local & 15,411 & 35,598 \\
Cufflinks & 25,318 & 34,959 \\
Global + Local + Cufflinks & 16,445 & 53,296\\
Ensembl & 17,934 & 23,392 \\
\hline
%table information
\end{tabular}
%\begin{flushleft}\footnotesize \textit{*Minumum set}
%\end{flushleft}
\label{genes_transcripts}

\caption{
\bf{Single-end reads mapped to gene models (maximum)}}
\begin{tabular}{cccccc}
\hline
Dataset & Mapped & Unmapped \\
\hline
Line 6 uninfected & 18,398,569 (78.03\%) & 5,180,983 (21.97\%) \\
Line 6 infected & 17,194,894 (73.33\%) & 6,254,087 (26.67\%) \\
Line 7 uninfected & 18,148,112 (75.85\%) & 5,777,697 (24.15\%) \\
Line 7 infected & 19,914,624 (78.52\%) & 5,447,943 (21.48\%) \\
\hline
%table information
\end{tabular}
%\begin{flushleft}Table Caption
%\end{flushleft}
\label{single-end_map}

\caption{
\bf{Paired-end reads mapped to gene models (maximum)}}
\begin{tabular}{cccccc}
\hline
Dataset & Mapped & Unmapped \\
\hline
Line 6 uninfected & 21,641,202 (77.65\%) & 12,022,675 (22.35\%) \\
Line 6 infected & 18,556,277 (77.61\%) & 5,352,216 (22.39\%) \\
Line 7 uninfected & 25,670,333 (77.97\%) & 7,250,999 (22.03\%) \\
Line 7 infected & 26,988,536 (78.27\%) & 7,489,299 (21.73\%) \\
\hline
%table information
\end{tabular}
%\begin{flushleft}Table Caption
%\end{flushleft}
\label{paired-end_map}
\end{table}

\section*{Figure Legends}
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{overall_pipeline.pdf}
\end{center}
\caption{
{\bf Gene model construction pipeline.} Transcripts are obtained from two assembly methods -- global and local assembly.
Transcripts are aligned to a chicken genome by BLAT. Gimme then constructs gene models based on alignments of transcripts.
}
\label{overall_pipeline}
\end{figure}

>>>>

We next used Tophat to align mRNAseq reads to the genome and partition
reads by chromosome; we then assembled the partitioned reads with
Velvet and Oases using the same range of parameters as the global
assembly, above.  While the local assemblies were considerably more
computationally efficient, they lacked several thousand unique regions
that were present in the global assembly
(Table~\ref{unique_sequences_matched_mouse}); this is probably due to
the incomplete nature of the current chicken genome (cite).
Interestingly, over a hundred regions were present {\em only} in the
local assemblies, suggesting that the local assemblies might be
recovering additional exons.  Significant numbers of unique regions
($>300$bp) from both global and local assemblies showed homology to
the mouse genome, indicating that at least some of these unique sequences
represented real sequence content.  Figure~\ref{global_vs_local} shows
an example of different isoforms detected by the two assembly methods.

@@

We developed an exon graph approach to construct gene models from
alignments of transcripts against the genome.  Our approach, called
Gimme, uses exon graphs to merge transcripts and gene models based on
overlapping exons (see Materials and Methods).  We used Gimme to
obtain 20,439 gene models containing 26,035 isoforms from our global
assembly; 21,963 gene models containing 25,859 from our local
assembly; and 20,922 gene models containing 27,732 isoforms from the
merged global and local assemblies (Table~\ref{genes_transcripts}).
The number of genes is slightly decreased after merging transcripts
from global and local assembly together. 
This indicates that some transcripts from both methods were merged together.
Figure~\ref{model_comparisons} shows gene models built by our method with
other existing gene models.

Most transcripts from RNA-Seq assembly include untranslated regions (UTRs),
especially $3'$ UTR in our datasets.
These regions are challenging to predict correctly solely from computational
methods due to low degree of conservation (cite).
RNA-Seq-derived gene models are useful for studying variations within UTR regions,
which may be involved in regulation of isoform expression (cite).
Figure~\ref{alter_5utr} and figure~\ref{long_utr} shows an example of alternative
$5'$UTRs and extended $3'$UTR respectively.

@@

\subsubsection*{Most splice junctions are independently supported}

@@

As shown above, Cufflinks and the assembly method detect a number of
distinct but equally valid splice junctions, which suggests that we
could obtain greater sensitivity by merging both sets of predictions.
We therefore used Gimme to merge mRNAs, the Cufflinks and assembly gene
models Table.~\ref{genes_transcripts}.  This resulted in an increased 
number of total genes and isoforms (e.g. see Fig.~\ref{mrna_cuff_gimme.pdf}).
The merged gene models recover 51\% and 66\% of splice junctions from ESTs
and ENSENBL respectively, which is 10-15\% greater than that from corresponding
unmerged gene models.
% @CTB what about splice junctions etc??

@@

\subsection*{Leftover text}

The program can merge all transcripts from multiple-kmers and predict
a structure of full-length isoforms and genes.
Total of 20,439 genes were obtained from global assembly and 21,963 genes
from local assembly.
However, the number of genes obtained from global + local assembly is only 20,922 genes.
The number of isoforms slightly increased after combining
transcripts from global and local assembly together (Table~\ref{genes_transcripts}).

Most transcripts include a large part of untranslated regions (UTRs),
especially $3'$ UTR in our datasets.
These regions are challenging to predict correctly solely from computational
methods due to low degree of conservation [?].
RNA-Seq-derived gene models are useful for studying variations within UTR
regions, which may be involved in regulation of isoform expression\cite{}

@@

The number of genes from our pipeline might be overestimated due to
fragmented transcripts.  In our datasets, read coverage is lower at
the $5'$ end due to the method used to convert mRNA to cDNA in library
preparation.  A bias of read coverage toward $3'$ end and low
expression level result in fragmented transcripts as shown in
Fig.~\ref{fragmented_transcripts}.  Moreover, spurious splice
junctions due to duplication in genome sequence can lead to a large
number of dubious isoforms.

@@

\subsubsection*{Cufflinks}

To evaluate an efficiency of our pipeline in detecting splice junctions,
we compared the number of splice junctions detected by our pipeline to
that from , ESTs and Ensembl gene models version 64.
Cufflinks detected 88,289 (87.93\%) splice junctions that are also detected
by our pipeline.
However, 42.43\% (5,143) of junctions not detected by Cufflinks are supported
by ESTs or mRNAs.
This suggests that the assembly detects some splice junctions that are
not detected by Cufflinks and they are genuine splice junctions.

We also aligned transcripts from Cufflinks and the pipeline to Ensembl
transcripts to compare genes and transcripts detected by both methods.
We selected only transcripts that match with more than 90\% identity and
cover more than 80\% of an Ensembl transcript.
Our pipeline detects 7,298 genes and Cufflinks detects 8,341 genes of
total 17,934 Ensembl genes.
6,798 of those genes are both detectd by our pipeline and Cufflinks.
In addition, we wanted to compare the number of isoforms detected by our
pipeline and Cufflinks within the same gene; however, a transcript can
match to multiple isoforms in the same gene, especially when it is not complete.
Therefore, we compare the number of splice junctions detected by both
methods within the same gene instead.
From 6,798 genes, our pipeline detects 56,332 splice junctions, whereas
Cufflinks detects 62,150 junctions.

However, because Ensembl gene models do not include all isoforms or
splice junctions from ESTs, we investigated splice junctions that are
not in Ensembl but are supported by ESTs in those 6,798 genes and
found that our pipeline detects 1,997 splice junctions and Cufflinks
only detects 1,860 splice junctions.
Figure ~\ref{alt_splice_site} shows an example of splice junctions not
included in Ensembl gene models, which are detected by our pipeline.

@@

Single and paired-end reads were mapped to chicken genome by Tophat2\cite{Trapnell:2009dp} release 2.0.0 using
default parameters without annotations.
All reads were mapped to cDNA sequences derived from gene models by Bowtie2\cite{Langmead:2009fv} with
default parameters \texttt{(n=2, l=28, e=70, k=1)}.
Reads from mouse dataset were mapped to mouse genome (mm9) downloaded from Tophat website \texttt{http://tophat.cbcb.umd.edu}.

@@

\caption{
\bf{Number of total and unique splice junctions}}
\begin{tabular}{cccccc}
\hline
Method& Total & Unique & Supported by ESTs \\ 
\hline
Oases-M assembly & 111,237 & 6,860 & 421 (6.1\%) \\
Unmerged assembly & 112,708 & 8,295 & 1,607 (19.4\%) \\
\hline
%table information
\end{tabular}
%\begin{flushleft}\footnotesize \textit{*Minumum set}
%\end{flushleft}
\label{Oases-M}

\caption{
\bf{Number of putative genes and isoforms}}
\begin{tabular}{cccccc}
\hline
Method& Gene & Isoform \\ 
\hline
Global & 20,439 & 26,035 \\
Local & 21,963 & 25,859 \\
Global + Local & 20,922 & 27,732 \\
Cufflinks & 25,318 & 34,959 \\
Global + Local + Cufflinks + mRNAs & 26,726 & 43,074\\
Ensembl & 17,934 & 23,392 \\
\hline
%table information
\end{tabular}
%\begin{flushleft}\footnotesize \textit{*Minumum set}
%\end{flushleft}
\label{genes_transcripts}

\caption{
\bf{Single-end reads mapped to gene models (maximum)}}
\begin{tabular}{cccccc}
\hline
Dataset & Mapped & Unmapped \\
\hline
Line 6 uninfected & 20,169,993 (85.54\%) & 3,409,559 (14.46\%) \\
Line 6 infected & 18,790,831 (80.13\%) & 4,658,150 (19.87\%) \\
Line 7 uninfected & 19,844,293 (82.94\%) & 4,081,516 (17.06\%) \\
Line 7 infected & 21,772,102 (85.84\%) & 3,590,465 (14.16\%) \\
\hline
%table information
\end{tabular}
%\begin{flushleft}Table Caption
%\end{flushleft}
\label{single-end_map}

\caption{
\bf{Paired-end reads mapped to gene models (maximum)}}
\begin{tabular}{cccccc}
\hline
Dataset & Mapped & Unmapped \\
\hline
Line 6 uninfected & 28,696,112 (85.36\%) & 4,967,765 (14.76\%) \\
Line 6 infected & 20,514,438 (85.93\%) & 3,394,055 (14.20\%) \\
Line 7 uninfected & 28,159,776 (85.65\%) & 4,761,556 (14.46\%) \\
Line 7 infected & 29,564,592 (85.88\%) & 4,913,243 (14.25\%) \\
\hline
%table information
\end{tabular}
%\begin{flushleft}Table Caption
%\end{flushleft}
\label{paired-end_map}
\end{table}

\section*{Figure Legends}
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{overall_pipeline.pdf}
\end{center}
\caption{
{\bf Gene model construction pipeline.} Transcripts are obtained from two assembly methods -- global and local assembly.
Transcripts are aligned to a chicken genome by BLAT\@. Gimme then constructs gene models based on alignments of transcripts.
}
\label{overall_pipeline}
\end{figure}
